The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-25 paper/paper-jsdse/paper-jsdse.tex
tweaking plots and captions

\begin{itemize}
\tightlist
\item
  \texttt{NLSY79.csv}: csv format
\item
  \texttt{NLSY79.dat}: dat format
\item
  \texttt{NLSY79.NLSY79}: Tagset of variables that can be uploaded to the website to recreate the data set.
\item
  \texttt{NLSY79.R}: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set, \texttt{categories\_qnames} (the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
\end{itemize}

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The input data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1980}, such as contain the information about the job number and the year. The input data consequently has a large number of columns (686 to be specific).

@@

\begin{verbatim}
#> Rows: 12,686
#> Columns: 11
#> $ CASEID_1979 <int> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1~
#> $ HRP1_1979   <int> 328, 385, 365, NA, 310, NA, NA, NA, 214, NA, 337~
#> $ HRP2_1979   <int> NA, NA, NA, NA, 375, NA, NA, NA, NA, NA, 300, NA~
#> $ HRP3_1979   <int> NA, NA, 275, NA, NA, NA, NA, NA, NA, NA, NA, NA,~
#> $ HRP4_1979   <int> NA, NA, NA, NA, NA, 250, NA, NA, NA, NA, NA, NA,~
#> $ HRP5_1979   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
#> $ HRP1_1980   <int> NA, 457, 397, NA, 333, 275, 300, 394, 200, 318, ~
#> $ HRP2_1980   <int> NA, NA, 367, NA, NA, NA, NA, NA, NA, NA, NA, NA,~
#> $ HRP3_1980   <int> NA, NA, 380, NA, NA, NA, 290, NA, NA, NA, NA, NA~
#> $ HRP4_1980   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
#> $ HRP5_1980   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
\end{verbatim}

\hypertarget{tidying-demographic-variables}{%
\subsubsection{Tidying demographic variables}\label{tidying-demographic-variables}}

@@

\caption{A glimpse of the refreshed data. Longitudinal profiles of wages from 1979-2018 are shown for a random sample of 36 individuals, divided into twelve subplots. Most individuals here experienced increasing wages over time, although there is considerable fluctuation from one year to another.}\label{fig:sample-plot}
\end{figure}

@@

{\centering \includegraphics[width=432px]{figures/feature-plot-1}

@@

{\centering \includegraphics[width=432px]{figures/compare-plot-1}

@@

\caption{Comparison between the original and the treated mean hourly wage. The orange line portray the original value of mean hourly wage, while the turquoise line display the mean hourly wages value after the extreme values imputed with the robust linear model's prediction value. We can see that some extreme spikes has been reduced by the model.}\label{fig:compare-plot}
\end{figure}

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value, transformed to log scale. The plot implies that the skewness is slightly negative. We also see that the three features are overlapped each other. It indicates that some ID's minimum wages are higher than some ID's maximum wages.

@@

{\centering \includegraphics[width=432px]{figures/fixed-feature-plot-1}

@@

{\centering \includegraphics[width=0.93\linewidth]{/Users/cookd/numbats/summer-wages-refresh/paper/figures/flowchart}

>>>>

\begin{itemize}
\tightlist
\item
  \texttt{NLSY79.csv}: csv format of the data
\item
  \texttt{NLSY79.dat}: dat format of the data
\item
  \texttt{NLSY79.NLSY79}: Tagset of variables that can be uploaded to the website to recreate the data set.
\item
  \texttt{NLSY79.R}: R script provided automatically by the database for reading the data into R and converting the variables' names and label into something more sensible. We utilized this code to do an initial data tidying. It produced two data set, \texttt{categories\_qnames} (the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
\end{itemize}

\hypertarget{tidywages}{%
\subsubsection{Tidying the wages data}\label{tidywages}}

The raw data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (686 to be specific). The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data output is shown below.

\begin{verbatim}
#> 'data.frame':    12686 obs. of  686 variables:
#>  $ CASEID_1979                  : int  1 2 3 4 5 6 7 8 9 10 ...
#>  $ HRP1_1979                    : int  328 385 365 NA 310 NA NA NA 214 NA ...
#>  $ HRP2_1979                    : int  NA NA NA NA 375 NA NA NA NA NA ...
#>  $ HRP3_1979                    : int  NA NA 275 NA NA NA NA NA NA NA ...
#>  $ HRP4_1979                    : int  NA NA NA NA NA 250 NA NA NA NA ...
#>  $ HRP5_1979                    : int  NA NA NA NA NA NA NA NA NA NA ...
#>  $ HRP1_1980                    : int  NA 457 397 NA 333 275 300 394 200 318 ...
#>  $ HRP2_1980                    : int  NA NA 367 NA NA NA NA NA NA NA ...
#>  $ HRP3_1980                    : int  NA NA 380 NA NA NA 290 NA NA NA ...
#>  $ HRP4_1980                    : int  NA NA NA NA NA NA NA NA NA NA ...
#>   [list output truncated]
\end{verbatim}

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data does not comply the rules of the tidy data, thus we wrangle the data in form required by making use of the \texttt{tidyverse} suite of packages (Wickham et al. 2019), which include the main set of packages needed, namely \texttt{tidyr} (Wickham 2020b), \texttt{dplyr} (Wickham et al. 2020), and \texttt{stringr} (Wickham 2019). The tidy data form, where we have a variable that indicates the individual ID, year, job number, wage in dollars and demographic variables, is important for downstream analysis, such as visualisation and modelling that will be performed later.

@@

\hypertarget{tidydemog}{%
\subsubsection{Tidying demographic variables}\label{tidydemog}}

@@

\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data. Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages. Some individuals are only measured for a short period.}\label{fig:sample-plot}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{figures/feature-plot-1}

@@

{\centering \includegraphics[width=1\linewidth]{figures/compare-plot-1}

@@

\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals. A robust linear model prediction was used to correct mean hourly wages value. We can see that some extreme spikes, corresponding to implausible wages, have been repleaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.}\label{fig:compare-plot}
\end{figure}

Further, \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value. We still can see some extreme values in maximum wages, but consider it as a natural variability of the data. In \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.

@@

{\centering \includegraphics[width=1\linewidth]{figures/fixed-feature-plot-1}

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/summer-wages-refresh/paper/figures/flowchart}

### 2021-09-25 paper/paper.Rmd
tweaking plots and captions

```{r sample-plot, echo = FALSE, fig.cap="A glimpse of the refreshed data. Longitudinal profiles of wages from 1979-2018 are shown for a random sample of 36 individuals, divided into twelve subplots. Most individuals here experienced increasing wages over time, although there is considerable fluctuation from one year to another.", fig.height=4, fig.width=5, out.width="100%"}
```

@@

```{r feature-plot, fig.cap = "Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C). Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.", fig.width=6, fig.height=2}
```

@@

```{r compare-plot, fig.cap="Comparison between the original and the treated mean hourly wage. The orange line portray the original value of mean hourly wage, while the turquoise line display the mean hourly wages value after the extreme values imputed with the robust linear model's prediction value. We can see that some extreme spikes has been reduced by the model.", fig.height=5, fig.width=6, outwidth="100%"}
```

@@

```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C). ", fig.width=6, fig.height=2}
```

@@

```{r flow-chart, fig.cap="The stages of data cleaning from the raw data to get three datasets contained yowie. n individuals means the number of respondents included in each stage, while n observations means the number of rows in the data. The color represented the stage of data cleaning in statistical value chain (van der Loo \\& de Jonge, 2018). Pink, blue, and green represents the raw, input, and valid data, respectively.", out.width="93%"}
```

>>>>

```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data. Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages. Some individuals are only measured for a short period.", fig.height=8, fig.width=10, out.width="100%"}
```

@@

```{r feature-plot, fig.cap = "Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C). Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.", fig.width=10, fig.height=4, out.width="100%"}
```

@@

```{r compare-plot, fig.cap="Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals. A robust linear model prediction was used to correct mean hourly wages value. We can see that some extreme spikes, corresponding to implausible wages, have been repleaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.", fig.height=8, fig.width=10, out.width="100%"}
```

@@

```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C). ", fig.width=10, fig.height=4, out.width="100%"}
```

@@

```{r flow-chart, fig.cap="The stages of data cleaning from the raw data to get three datasets contained yowie. n individuals means the number of respondents included in each stage, while n observations means the number of rows in the data. The color represented the stage of data cleaning in statistical value chain (van der Loo \\& de Jonge, 2018). Pink, blue, and green represents the raw, input, and valid data, respectively.", out.width="85%"}
```
